Kindly Bent to Free Us

### 2020-06-20 kindly/soa.tex
matrix and explanation

    \begin{tabular}{l|ccccccccc}
      \textbf{Language}
      & UAL & \rot{Borrows} & \rot{State}
      & \rot{Resource Subsumption}
      & \rot{Resource Polymorphism}
      & \rot{Identity}
      & \rot{Inference} & \rot{Formalisation}
      & \rot{Basis}
      \\\hline
      System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10}
      & UL & --- & \N & \N & \N & \N & \N & \YC Coq & System F
      \\
      Alms~\citep{DBLP:conf/popl/TovP11}
      & UA & --- & \Y & \M & \Y & \Y & \MC Local & \N & ML \\
      Quill~\citep{DBLP:conf/icfp/Morris16}
      & UL & --- & \N & \N & \Y & \N & \YC Principal & \MC Manual & Qualified types\\
      Lin. Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
      & UL & --- & \M & \N & \Y & \N & \MC Non-principal & \MC Manual & Haskell \\
      Mezzo~\citep{DBLP:phd/hal/Protzenko14}
      & UA & --- & \Y & \M & \Y & \Y & \MC Local & \YC Coq & --- \\
      \hline
      Rust~\citep{rust,DBLP:journals/pacmpl/0002JKD18}
      & UA & \Y & \Y & \Y & \M & \N & \MC Local & \YC Coq & --- \\
      Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09}
      & UA & --- & \Y & \Y & \Y & \Y & \N & \MC Manual & Java \\
      \hline
      \lang
      & UAL & \Y & \Y & \Y & \Y & \N & \YC Principal & \MC Manual & ML/HM(X)
      \\
    \end{tabular}
  \end{center}
  \caption{Comparison matrix}
  \label{fig:comparison-matrix}
\end{figure}

@@

The comparison matrix in \cref{fig:comparison-matrix} gives an
overview over the systems discussed in this section. The column UAL
specifies the substructural features supported (unrestricted, affine,
linear). Borrows and state indicate support for the respective
feature. ``Resource Subsumption'' indicates that unrestricted elements 
can be promoted to affine and then linear. This promotion applies to
all objects, resources,
borrows, or closures. ``Resource Polymorphism'' refers to polymorphism over
substructural features, notably for closures.
``Identity'' indicates that the language supports a notion
of identity, usually through existential types, as described in
\cref{identity}.
Inference is self-explanatory, formalization
refers to the existence of a formal semantics and type soundness proof. Basis is self-explanatory.

@@

Mezzo~\citep{DBLP:phd/hal/Protzenko14} is an ML-like language
with a rich capability system which is able to encode numerous
properties akin to separation logic~\citep{DBLP:conf/lics/Reynolds02}.
Mezzo explores the  boundaries of the design space of type systems for
resources. Hence, it is more expressive than \lang, but
much harder to use. The Mezzo typechecker relies on explicit
annotations and it is not known whether type inference for Mezzo is possible.

@@

Vault~\citep{DBLP:conf/pldi/DeLineF01}
and Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09}
leverage typestate and capabilities
to express rich properties in objects and protocols.
These systems are designed for either low-level or object-oriented
programming and do not immediately lend themselves to a more functional
style. While these systems are much more
powerful than \affe's, they require programmer annotations
and do not support inference.
It  would be interesting to extend \lang with limited
forms of typestate as a local, opt-in feature to provide
more expressivity at the cost of inference.

>>>>

    \begin{tabular}{l|c@{~}ccccccccc}
      \textbf{Language}
      & UAL & \rot{Borrows} & \rot{State}
      & \rot{Resource Subsumption}
      & \rot{Resource Polymorphism}
      & \rot{Identity}
      & \rot{Concurrency}
      & \rot{Inference} & \rot{Formalisation}
      & \rot{Basis}
      \\\hline
      System F\degree~\citep{DBLP:conf/tldi/MazurakZZ10}
      & UL & \N & \N & \N & \N & \N & \N & \N & \YC Coq & System F
      \\
      Alms~\citep{DBLP:conf/popl/TovP11}
      & UA & \N & \Y & \M & \Y & \Y & \Y & \MC Local & \N & ML \\
      Quill~\citep{DBLP:conf/icfp/Morris16}
      & UL & \N & \N & \N & \Y & \N & \N & \YC Principal & \MC Manual & Qualified types\\
      Lin. Haskell~\citep{DBLP:journals/pacmpl/BernardyBNJS18}
      & UL & \N & \M & \N & \Y & \N & \M & \MC Non-pr. & \MC Manual & Haskell \\
      Mezzo~\citep{DBLP:phd/hal/Protzenko14,DBLP:journals/toplas/BalabonskiPP16}
      & UA & \M & \Y & \M & \Y & \Y & \Y & \MC Local & \YC Coq & ML \\
      \hline
      Rust~\citep{rust,DBLP:journals/pacmpl/0002JKD18}
      & UA & \Y & \Y & \Y & \M & \N & \Y & \MC Local & \YC Coq & --- \\
      Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09,DBLP:journals/toplas/GarciaTWA14}
      & UA & \N & \Y & \Y & \Y & \Y & \N   & \N & \MC Manual & Java \\
      \hline
      \lang
      & UAL & \Y & \Y & \Y & \Y & \N & \N & \YC Principal & \MC Manual & ML/HM(X)
      \\
    \end{tabular}
  \end{center}
  \caption{Comparison matrix}
  \label{fig:comparison-matrix}
\end{figure}

@@

The comparison matrix in \cref{fig:comparison-matrix} gives an
overview over the systems discussed in this section.
Each column indicates whether a feature is present (\Y), absent (\N), or
partially supported (\M), i.e., if the feature is limited or can
only be obtained through a non-trivial encoding.
Features are selected according to their contribution to expressivity
and programmer convenience.
The column UAL
specifies the substructural features supported (unrestricted, affine,
linear). Borrows and state indicate support for the respective
feature. ``Resource Subsumption'' indicates that unrestricted elements 
can be promoted to affine and then linear. This promotion applies to
all objects, resources,
borrows, or closures. ``Resource Polymorphism'' refers to polymorphism over
substructural features, notably for closures.
``Identity'' indicates that the language supports a notion
of identity, usually through existential types, as described in
\cref{identity}.
``Concurrency'' indicates whether the language supports
concurrency. For example, the implementation of Linear Haskell
supports state and concurrency, but its theory covers neither.
Type ``inference''  can be local, principal, or non-principal (if the
inferred type is not necessary the most general). ``Formalization''
refers to the existence of a formal semantics and type soundness
proof. ``Basis'' indicates the heritage or inspiration of the language.

@@

Mezzo~\citep{DBLP:phd/hal/Protzenko14,DBLP:journals/toplas/BalabonskiPP16} is an ML-like language
with a rich capability system which is able to encode numerous
properties akin to separation logic~\citep{DBLP:conf/lics/Reynolds02}.
Mezzo explores the  boundaries of the design space of type systems for
resources. Hence, it is more expressive than \lang, but
much harder to use. The Mezzo typechecker relies on explicit
annotations and it is not known whether type inference for Mezzo is possible.

@@

Vault~\citep{DBLP:conf/pldi/DeLineF01}
and Plaid~\citep{DBLP:conf/oopsla/AldrichSSS09,DBLP:journals/toplas/GarciaTWA14}
leverage typestate and capabilities
to express rich properties in objects and protocols.
These systems are designed for either low-level or object-oriented
programming and do not immediately lend themselves to a more functional
style. While these systems are much more
powerful than \affe's, they require programmer annotations
and do not support inference.
It  would be interesting to extend \lang with limited
forms of typestate as a local, opt-in feature to provide
more expressivity at the cost of inference.
